\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}	% Para caracteres en español
\usepackage{amsmath,amsthm,amsfonts,amssymb,amscd}
\usepackage{multirow,booktabs}
\usepackage[table]{xcolor}
\usepackage{fullpage}
\usepackage{lastpage}
\usepackage{enumitem}
\usepackage{fancyhdr}
\usepackage{mathrsfs}
\usepackage{wrapfig}
\usepackage{setspace}
\usepackage{siunitx}
\usepackage{calc}
\usepackage{multicol}
\usepackage{cancel}
\usepackage[retainorgcmds]{IEEEtrantools}
\usepackage[margin=3cm]{geometry}
\usepackage{amsmath}
\newlength{\tabcont}
\setlength{\parindent}{0.0in}
\setlength{\parskip}{0.05in}
\usepackage{physics}
\usepackage{empheq}
\usepackage{framed}
\usepackage[most]{tcolorbox}
\usepackage{xcolor}

\usepackage{pgfplots}
\pgfplotsset{width=10cm,compat=1.9}
\usepgfplotslibrary{external}
\tikzexternalize

\colorlet{shadecolor}{orange!15}
\parindent 0in
\parskip 12pt
\geometry{margin=1in, headsep=0.25in}
\theoremstyle{definition}
\newtheorem{defn}{Definition}
\newtheorem{reg}{Rule}
\newtheorem{exer}{Exercise}
\newtheorem{note}{Note}
\begin{document}
\setcounter{section}{8}
\title{Chapter 9 Review Notes}

\thispagestyle{empty}

\begin{center}
{\LARGE \bf Quantum Chemistry}\\
{\large CML101}\\
Salil Gokhale
\end{center}

\section{Postulates of Quantum Mechanics}
\subsection{Postulate I: The Wavefunction}

\textbf{The state of the system is described as fully as possible by the wavefunction $\psi(r_1, r_2,\dots ,t)$ where $r_1$, $r_2, \dots$ are the locations of the particles and $t$ is time.}

$\psi(r_1, r_2,\dots ,t) \rightarrow$ time-dependent wavefunction

$\psi(r_1, r_2,\dots) \rightarrow$ time-independent wavefunction

\subsection{Postulate II: The Born Interpretation} 
\textbf{For a system described by the wavefunction $\psi(r_1, r_2, \dots)$ the probability of finding particle 1 in the volume element $\dd{\tau_1}$ at $r_1$, particle 2 in the volume element $\dd{\tau_2}$ at $r_2$, etc. is proportional to $\abs{\psi}^2 \dd{\tau_1} \dd{\tau_2} \dots$}

$\abs{\psi}^2 \rightarrow$ probability density

$\psi \rightarrow$ probability amplitude. It has no physical significance 

\subsubsection{Normalization Condition} 

A wavefunction is said to be normalized if $\int \psi^* \psi \dd{\tau} =1$

Hence the normalisation constant $N$(which when multiplied to the wavefunction normalises it) is equal to $\frac{1}{\left(\int \psi^* \psi \dd{\tau} \right)^{1/2}}$. If this integral has a finite value, the wavefunction is \textbf{square-integrable}.

\subsection{Constraints on Wavefunction}

\begin{itemize}
    \item $\psi$ must not be infinite over a finite region
    \item $\psi$ must be single-valued
    \item $\psi$ must be continuous
    \item $\dv{\psi}{x}$ must be continuous
\end{itemize}

\subsection{Postulate III: Quantum Mechanical Operators}

\textbf{For each observable property $\Omega$ of a system there is a corresponding operator $\vectorunit{\Omega}$ built from the following position and momentum operators:}

\begin{itemize}
    \item $\vectorunit{x} = x \times$
    \item $\vectorunit{p} = -\frac{\hbar}{i} \grad$ or in special case of one dimension $\vectorunit{p_x} = -\frac{\hbar}{i} \dv{}{x}$
\end{itemize}

All the operators we encounter are linear i.e. $\vectorunit{\Omega}(\psi_1 + \psi_2) =  \vectorunit{\Omega}(\psi_1) + \vectorunit{\Omega}(\psi_2)$

\begin{shaded}
\textbf{Time-independent Schrodinger Equation}
\begin{equation*}
\vectorunit{H}\psi = E\psi
\end{equation*}

where,

\begin{gather*}
\vectorunit{H} = \frac{-\hbar^2}{2m} \laplacian + V \text{ is the hamiltonian operator corresponding to total energy of system.} \\
E \text{ is the total energy of system}
\end{gather*}
\end{shaded}



This suggests the form of the \textbf{eigenvalue equation} 

\begin{equation*}
    \vectorunit{\Omega} \psi = \omega \psi
\end{equation*}

where,
\begin{gather*}
    \vectorunit{\Omega} \text{ is a operator corresponding to an observable} \\
    \omega\text{ is called observable or \textbf{eigenvalue} of operator. It is always a real value.}\\
    \psi \text{ is the wavefunction and also called as \textbf{eigenfunction} of the operator}
\end{gather*}
\subsection{Postulate IV: Eigenvalues and Eigenfunctions}
\textbf{If the system is described by a wavefunction $\psi$ that is an eigenfunction of $\vectorunit{\Omega}$ such that $\vectorunit{\Omega} \psi = \omega \psi$ then the outcome of a measurement of $\Omega$ will be the eigenvalue $\omega$}

eigenvalue is real $\iff$ the operator $\vectorunit{\Omega}$ is Hermitian:

\begin{equation*}
    \int \psi_1 ^* \vectorunit{\Omega} \psi_2 \dd{x} = \left[ \int \psi_2 ^* \vectorunit{\Omega} \psi_1 \right]^*
\end{equation*}
The momentum and position operators are hermitian. Sum and product of two hermitian operators is also hermitian.

\subsection{Postulate V: Predicting Experiments}
\textbf{When the value of an observable $\Omega$ is measured for a system that is described by a wavefunction which is a linear combination, $\psi = \sum_k c_k \psi_k$, each measurement of $\Omega$ gives one of the eigenvalues $\omega_k$ with a probability proportional to $\abs{c_k}^2$.}
If $\psi$ is normalized, then this probability is equal to $\abs{c_k}^2$.


\begin{shaded}
The mean of the measurements of $\Omega$ is equal to expectation value of the operator $\vectorunit{\Omega}$

\begin{equation*}
    \expval{\Omega} = \frac{\int \psi^* \vectorunit{\Omega} \psi \dd{\tau}}{\int \psi^* \psi \dd{\tau}}
\end{equation*}
for a normalized wavefunction, the denominator is unity. If $psi$ is an eigenfunction of $\vectorunit{Omega}$ then each measurment and consequently expectation value will be equal to it's eigenvalue $\omega$.

In the case of $\psi = \sum_{k} c_k \psi_k$, the expectation value is($\psi$ is normalized)
\begin{equation*}
    \expval{\Omega} = \sum_{k} \abs{c_k}^2 \omega_k
\end{equation*}

\end{shaded}
\subsubsection{Orthogonality of Eigenfunctions}
\textbf{Eigenfunctions corresponding to different eigenvalues of the same Hermitian operator are orthogonal.} To say that two different functions $\psi_i$ and $\psi_j$ are orthogonal means that the integral (over all space) of their product is zero:

\begin{equation*}
    \int \psi_i ^* \psi_j \dd{\tau} = 0
\end{equation*}
in a special case when both functions are normalised,

\begin{equation*}
    \int \psi_i ^* \psi_j \dd{\tau} = \delta_{ij}
\end{equation*}

\section{The Uncertainty Principle}
\begin{shaded}
\textbf{Heisenberg Uncertainty Principle}
\begin{equation*}
    \Delta p_x \Delta x \geq \frac{\hbar}{2}
\end{equation*}
where,
\begin{gather*}
     \Delta X = \left[\expval{X^2} - \expval{X}^2 \right]^{1/2}
\end{gather*}

Note that there is no such constraint on say, $p_y$ and $x$
\end{shaded}

\subsection{General Form of Uncertainty principle}

Two observables $\Omega_1$ and $\Omega_2$ are complementary if $\vectorunit{\Omega}_1 \vectorunit{\Omega}_2 \psi \neq  \vectorunit{\Omega}_2 \vectorunit{\Omega}_1 \psi$

The commutator is of the two operators is defined as their difference: $\comm{\vectorunit{\Omega}_1}{\vectorunit{\Omega}_2} = \vectorunit{\Omega}_1\vectorunit{\Omega}_2 - \vectorunit{\Omega}_2\vectorunit{\Omega}_1$. If two operators commute then their commutator is 0.

\begin{shaded}
\textbf{General Uncertainty Principle}
\begin{equation*}
    (\Delta \Omega_1)^2 (\Delta \Omega_2)^2 \geq \frac{1}{4} \expval{i \comm{\vectorunit{\Omega}_1}{\vectorunit{\Omega}_2}}^2
\end{equation*}
from which it immediately follows that
\begin{equation*}
    \Delta \Omega_1 \Delta \Omega_2 \geq \frac{1}{2} \abs{\expval{i \comm{\vectorunit{\Omega}_1}{\vectorunit{\Omega}_2}}}
\end{equation*}
\end{shaded}




\section{Motion in 1D}

A particle with net motion is described by a complex wavefunction. A wavefunction that is purely real corresponds to zero net motion.

For a freely moving particle in one dimension, schrody's equation is

\begin{shaded}
\begin{equation*}
    - \frac{\hbar}{2m} \dv[2]{\psi(x)}{x} = E \psi(x)
\end{equation*}

Solving this we get the general solutions as,

\begin{gather*}
    \psi_k = Ae^{ikx} + Be^{-ikx}\\
    E_k = \frac{k^2 \hbar^2}{2m}
\end{gather*}
\end{shaded}

It follows that in this case, energy is \textbf{not} quantized.

\subsection{Particle in a 1D box}

Particle is confined between $x=0$ and $x=L$. Hence potential energy is:
\begin{equation*}
V=\begin{cases}
          0 \quad &\text{if } \, x \in (0,L) \\
          \infty \quad &\text{if } \, x \in (-\infty,0] \cup [L, \infty) \\
     \end{cases}
\end{equation*}

For particle inside walls, wavefunction will be same as that for free motion. It would be more convenient to write $\psi_k =Ae^{ikx} + Be^{-ikx}$ as $\psi_k = C\sin{kx} + D \cos{kx}$ where $C = (A-B)i$, $D = (A+B)$. For particle outside walls, wavefunction is 0.

\begin{equation*}
\psi_k(x) =\begin{cases}
          C\sin{kx} + D \cos{kx} \quad &\text{if } \, x \in (0,L) \\
          0 \quad &\text{if } \, x \in (-\infty,0] \cup [L, \infty) \\
     \end{cases}
\end{equation*}
At the moment, it seems that $k$ can take any value.

\subsubsection{Restricting values of $k$}

For continuity of wavefunction, it must satisfy the following boundary conditions:
\begin{itemize}
    \item $\psi_k (0) = 0$
    \item $\psi_k (L) = 0$
\end{itemize}

First condition implies that $D=0$ and second condition implies $k = \frac{n\pi}{L}$. Normalizing the wavefunction gives $C = \left(\frac{2}{L}\right)^{1/2}$. This will quantize the system by allowing only discrete values.
\begin{shaded}
\begin{equation*}
\psi_n(x) =\begin{cases}
          \left(\frac{2}{L}\right)^{1/2}\sin{\frac{n\pi x}{L}} \quad &\text{if } \, x \in (0,L) \\
          0 \quad &\text{if } \, x \in (-\infty,0] \cup [L, \infty) \\
     \end{cases}
\end{equation*}
and 
\begin{equation*}
    E_n = \frac{n^2 \pi^2 \hbar^2}{2mL^2} = \frac{n^2 h^2}{8mL^2} \quad  \text{ where }n=1,2,3,\dots
\end{equation*}
\end{shaded}
\subsubsection{Properties of Wavefunction}
\textbf{Nodes are points where the wavefunction passes through zero (not merely reaching zero, as at the walls)}. This particular wavefunction has $n-1$ nodes.

\begin{tikzpicture}
\begin{axis}[
    axis lines = center,
    ylabel = {\(\psi(x)\)},
    yticklabels={,,},
    xticklabels={,,},
    legend pos=outer north east,
]

\addplot [
    domain=0:pi, 
    samples=100, 
    color=red,
]
{sin(deg(x))};
\addlegendentry{\(n=1\)}

\addplot [
    domain=0:pi, 
    samples=100, 
    color=blue,
]
{sin(deg(2*x))};
\addlegendentry{\(n=2\)}

\addplot [
    domain=0:pi, 
    samples=100, 
    color=green,
]
{sin(deg(3*x))};
\addlegendentry{\(n=3\)}

\addplot [
    domain=0:pi, 
    samples=100, 
    color=yellow,
]
{sin(deg(4*x))};
\addlegendentry{\(n=4\)}

\addplot[mark=*] coordinates {(pi,0)} node[pin=100:{$L$}]{} ;

\end{axis}
\end{tikzpicture}

The probability density $\psi^2(x)$ becomes more uniform as $n$ increases, provided we ignore the fine detail of the increasingly rapid oscillations. The probability density at high quantum numbers reflects the classical result that a particle bouncing between the walls spends, on the average, equal times at all points. Thus classical mechanics emerges from quantum mechanics as high quantum numbers are reached.
\subsubsection{Properties of Observables}

$$\psi_n(x) = \left(\frac{2}{L}\right)^{1/2}\sin{\frac{n\pi x}{L}} = \frac{1}{2i}\left(\frac{2}{L}\right)^{1/2} (e^{ikx} + e^{-ikx})$$
It follows that linear momentum will be $k\hbar$ for half measurments and $-k\hbar$ for other half. So expectation value is 0. Which we could even obtain directly by integration:

\begin{equation*}
    \expval{p} =  \frac{2\hbar n \pi}{iL^2} \int_0 ^L \sin{\frac{n\pi x}{L}} \cos{\frac{n \pi x}{L}} \dd{x} = 0
\end{equation*}

Because $n$ cannot be zero, the lowest energy that the particle may possess is not zero (as would be allowed by classical mechanics) but this lowest, irremovable energy is called the \textbf{zero-point energy}. The physical origin of the zero-point energy can be explained in two ways:

\begin{itemize}
    \item The Heisenberg uncertainty principle requires a particle to possess kinetic energy if it is confined to a finite region(which in this case is true)
    \item If the wavefunction is to be zero at the walls, but smooth, continuous, and not zero everywhere, then it must be curved, and curvature in a wavefunction implies the possession of kinetic energy.
\end{itemize}
\section{Motion in 2D and 3D}
\subsection{Motion in 2D box}
Let the region of box be $\{(x,y) \mid 0<x<L_1, 0<y<L_2 \}$
Schrody's equation will be 

\begin{equation*}
    \frac{-\hbar^2}{2m} \left(\dv[2]{\psi}{x} + \dv[2]{\psi}{y}\right) = E \psi
\end{equation*}

\subsubsection{Separation of Variables}

Assume that $\psi$ can be written as a product of two function $X(x)$ and $Y(y)$. I.e. $\psi(x,y) = X(x) Y(y)$.

Substituting this in schrody's equation, dividing both sides by $XY$ and rearranging,
\begin{equation*}
    \frac{1}{X}\dv[2]{X}{x} + \frac{1}{Y} \dv[2]{Y}{y} = -\frac{2mE}{\hbar^2}
\end{equation*}
The first term on the left is independent of $y$, so if $y$ is varied only the second term of the two on the left can change. But the sum of these two terms is a constant given by the right-hand side of the equation. Hence both the terms of left are constants. We can then break it like this:

\begin{gather*}
    \frac{1}{X}\dv[2]{X}{x} =-\frac{2mE_X}{\hbar^2} \quad \text{and} \quad \frac{1}{Y}\dv[2]{Y}{y} =-\frac{2mE_Y}{\hbar^2}\\
    -\frac{\hbar^2}{2m} \dv[2]{X}{x} = E_X X \quad \text{and} \quad  -\frac{\hbar^2}{2m} \dv[2]{Y}{y} = E_Y Y\\
    \text{where, }\\
    E = E_X + E_Y
\end{gather*}

This is exactly the equation for a particle in 1D box. So we can directly write the wavefunction easily:

\begin{shaded}
\textbf{2D box}
\begin{gather*}
    \psi_{n_1, n_2}(x,y) = \begin{cases}
    \frac{2}{(L_1 L_2)^{1/2}} \sin{\frac{n_1 \pi x}{L_1}} \sin{\frac{n_2 \pi y}{L_2}} \quad \text{for} \quad 0<x<L_1, 0<y<L_2\\
    \\
    0 \quad \text{otherwise}
    \end{cases}
\end{gather*}

Simillarly, $E = E_x + E_y$
\begin{equation*}
    E_{n_1, n_2} = \frac{h^2}{8m} \left(\frac{n_1^2}{L_1^2} + \frac{n_2^2}{L_2^2}\right)
\end{equation*}
\end{shaded}

\subsubsection{Degeneracy}

Now consider a situation where $L_1 = L_2 = L$. Then $n_1 = 1, n_2 = 2$ and $n_2 = 1, n_1 = 2$ will yield different wavefunctions but same energy. The technical term for different wavefunctions with same energy is \textbf{degeneracy.} In this case, we say the state is \textbf{double degenerate}.

Each occurrence of degeneracy is related to a symmetry of the system. In this case it's rotating the box by $90^0$ changes $\psi_{1,2}$ into $\psi_{2,1}$.
\subsection{Motion in 3D box}

Using similar separation of variables we get the following:
\begin{shaded}
\textbf{3D box}
\begin{gather*}
    \psi_{n_1, n_2, n_3}(x,y,z) = \begin{cases}
    \frac{2\sqrt{2}}{(L_1 L_2 L_3)^{1/2}} \sin{\frac{n_1 \pi x}{L_1}} \sin{\frac{n_2 \pi y}{L_2}} \sin{\frac{n_3 \pi z}{L_3}} \quad \text{for} \quad 0<x<L_1, 0<y<L_2, 0<z<L_3\\
    \\
    0 \quad \text{otherwise}
    \end{cases}
\end{gather*}

Simillarly, $E = E_x + E_y + E_z$
\begin{equation*}
    E_{n_1, n_2,n_3} = \frac{h^2}{8m} \left(\frac{n_1^2}{L_1^2} + \frac{n_2^2}{L_2^2} + \frac{n_3^2}{L_3^2d}\right)
\end{equation*}
\end{shaded}


\section{Dirac Notation}

\begin{equation*}
    \ket{\psi} = \sum_{i=0}^n a_{i}\ket{\vectorunit{\psi}_i}
\end{equation*}
This is called a \textbf{ket}. It is a column vector with coefficients $a_i$(which may be complex) along the unit basis vectors $\ket{\vectorunit{\psi}_i}$. We can write $\ket{\psi}$ as a column vector:
\begin{equation}
   \ket{\psi}= \mqty[a_1 \\ a_2 \\ \vdots \\ a_n]
\end{equation}

A \textbf{bra} is a row vector which consists of complex conjugates of $a_i$. It is shows as $\bra{\psi}$

\begin{equation}
   \bra{\psi}= \mqty[a_1^* & a_2^* & \dots & a_n^*]
\end{equation}

Scalar product of $\bra{\psi_i}$ and $\ket{\psi_j}$ is just written as $\bra{\psi_i}\ket{\psi_j} = \sum_{i=0}^n a_i^* b_i$. The order of scalar product matters if the coefficients are complex numbers because $\bra{\psi_i}\ket{\psi_j} =  \bra{\psi_j}\ket{\psi_i}^*$

It follows that if the wavefunctions are normalised,  $\bra{\psi_i}\ket{\psi_j} = \delta_{ij}$

\textbf{Because $\bra{\psi}$ and $\ket{\psi}$ exist in different mathematical spaces, we cannot add or subtract them.}

\subsection{Observables and Operators}

The eigenvalue equation can be written as:

\begin{equation*}
    \vectorunit{\Omega} \ket{\psi} = \omega \ket{\psi}
\end{equation*}
where $\psi$ is an \textbf{eigenvector} of $\vectorunit{\Omega}$.

Multiplying both sides by $\bra{\psi}$ gives

\begin{equation*}
    \ev{\vectorunit{\Omega}}{\psi} = \ev{\omega}{\psi} = \omega \bra{\psi}\ket{\psi}
\end{equation*}
because $\omega$ is just a scalar real value. And if $\bra{\psi}$ is normalised, we can easily write the expectation value of $\omega$ in state $\psi$.

\begin{shaded}
\textbf{Expectation Value}
\begin{equation*}
    \expval{\omega}_{\psi} =  \ev{\vectorunit{\Omega}}{\psi} = \int \psi^* \vectorunit{\Omega} \psi \dd{\tau}
\end{equation*}
This is the derivation for expectation value in Postulate V.
\end{shaded}

Properties of operators (some already discussed before):

\begin{itemize}
    \item Operators are linear $\vectorunit{\Omega} (\ket{\psi_1} + \ket{\psi_2}) =\vectorunit{\Omega}  \ket{\psi_1} + \vectorunit{\Omega}\ket{\psi_1} $
    \item Operators with real eigenvalues are Hermitian $\bra{\psi_1}\vectorunit{\Omega}\ket{\psi_2} = \bra{\psi_2}\vectorunit{\Omega}\ket{\psi_1}$
    \item Hermitian operator with a set of eigenvectors which have different eigenvalues are orthogonal $\bra{\psi_i}\ket{\psi_j} = 0$
    \item There exists a set of eigenvectors of both operators $\vectorunit{\Omega}_1$ and $\vectorunit{\Omega}_2$ $\iff$ these operators commute, $\comm{\vectorunit{\Omega}_1}{\vectorunit{\Omega}_2} = 0$
    \begin{shaded}
    \textbf{Proof}
    \begin{gather*}
        \vectorunit{\Omega}_1 \ket{\psi} = \omega_1 \ket{\psi}\\
        \vectorunit{\Omega}_2 \ket{\psi} = \omega_2 \ket{\psi}
    \end{gather*}
    Multiplying first equation by $\vectorunit{\Omega}_2$ and second equation by $\vectorunit{\Omega}_1$ gives,
    \begin{gather*}
        \vectorunit{\Omega}_2 \vectorunit{\Omega}_1 \ket{\psi} = \omega_1 \vectorunit{\Omega}_2 \ket{\psi} = \omega_1 \omega_2 \ket{\psi}\\
        \vectorunit{\Omega}_1 \vectorunit{\Omega}_2 \ket{\psi} = \omega_2 \vectorunit{\Omega}_1 \ket{\psi} = \omega_2 \omega_1 \ket{\psi}\\
    \end{gather*}
    
    Subtracting them gives the required result.
    \end{shaded}
\end{itemize}

    


\end{document}